Researchers preferentially collaborate with same-gendered colleagues across the life sciences

### 2018-06-05 manuscript/Manuscript_Claire.Rmd
Many changes after Devi comments

---
title: "Do researchers preferentially collaborate with colleagues of the same gender?"
author:
- Luke Holman* and Claire Morandin\textsuperscript{§}
- '*\textit{luke.holman@unimelb.edu.au}'
- \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}
bibliography: references.bib
output:
  pdf_document: default
  html_document: default
  word_document: default
fontsize: 12pt
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Gender and coauthorship} \usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers
csl: plos.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious
  biases, and often feel excluded from formal and informal opportunities for research
  collaboration. In addition to ensuring fairness and helping to redress gender imbalance
  in the academic workforce, increasing women's access to collaboration could help
  scientific progress by drawing on more of the available human capital. Here, we
  test whether researchers preferentially collaborate with same-gendered colleagues,
  using more stringent methods and a larger dataset than in past work. Our results
  reaffirm that researchers co-publish with colleagues of the same gender, and show
  that this 'gender homophily' is slightly stronger today than it was 10 years ago.
  Contrary to our expectations, we found no evidence that homophily is driven mostly
  by senior academics, and no evidence that homophily is strongest in fields where
  women are in the minority. Interestingly, journals with a high impact factor for their discipline tended to have comparatively low homophily, as predicted if mixed-gender teams produce better research.
  We discuss potential causes of gender homophily in science. \vspace{5mm} \par\noindent
  \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining,
  Women in STEM. 
---

@@

# Load some little files with numbers to be inserted into the MS
sample.size <- read.csv("sample.size.table.csv", stringsAsFactors = FALSE)
sample.size[,2] <- prettyNum(sample.size[,2], big.mark = ",")
sample.size[,3] <- prettyNum(sample.size[,3], big.mark = ",")
names(sample.size)[2:3] <- c("n2005", "n2015")

@@

R2 <- readRDS("R2_for_figure4.RData") # For Fig 4
```

@@

Because publishing, networking and collaboration are instrumental to scientific productivity and academic career advancement [@Lee_2005; @Wuchty_2007; @Abramo_2009; @Lariviere_2015], dozens of studies have tested for gender differences in these areas [e.g. @Long_1992; @Bozeman_2011; @Abramo_2013; @Badar_2013; references in Table S1 of @Holman_2018]. For example, studies have found that women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less within their own university departments [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in total [@Fuchs_2001]. These gender differences in collaboration practice presumably have multiple causes, which might include implicit and explicit gender bias [@Moss_2012], differential family obligations [@Reskin_1978; @Long_1990; @Wright_2003], gender differences in confidence or self-esteem [@bleidorn2016age], concerns relating to sexual harassment [@jagsi_2016], and unequal access to conferences [@Martin_2014] and travel funds [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008]. Additionally, empirical studies imply that mixed-gender or otherwise diverse teams produce better results on collaborative tasks than less diverse teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. For reasons such as these, several studies have examined the author lists of published research articles in order to test for gender differences in collaboration frequency or pattern. To our knowledge, all of these studies imply that men co-publish with other men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing is often termed 'gender homophily'.

@@

![The Wahlund effect can make it appear as if authors prefer to publish with same-gendered colleagues, even if no such preference exists. Here, coloured circles represent male and female authors, and coauthors are linked with lines. Across the whole set of ten papers, there is an apparent excess of same-gender collaborations: there are six same-gender papers and only four mixed-gender papers, which is fewer than the $10\times2\times0.5\times0.5 = 5$ mixed-gender papers expected under the null hypothesis that authors assort randomly with respect to gender. However, within each subset, there is no evidence that authors prefer to publish with same-gendered individuals (if anything, this small dataset suggests gender heterophily). The Wahlund effect will tend to inflate the frequency of same-sex coauthorships whenever the data is composed of two or more disconnected subsets of literature with different author gender ratios; these subsets could be research disciplines, older versus newer papers, or papers from authors in different countries. \label{wahlund_plot}](../figures/Figure 1 - Wahlund figure.pdf)

@@

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. The great majority of journals had $\alpha' > 0$, and for many of these the FDR-corrected p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6; S1 Data). Only 2/2077 journals had statistically significantly heterophily (i.e. $\alpha' < 0$) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, consistent with the null hypothesis of random assortment with respect to gender.

![Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of gender homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after false discovery rate correction. The white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. \label{alpha_histograms}](../figures/figure2.pdf)

$\alpha'$ was significantly higher in the literature sample from 2015-16 relative to 2005-6, though the difference in means was small (S1 Fig; Effect of the fixed factor 'Time period' in a linear mixed model of the data for all author positions: Cohen's $d$ = $0.091{\pm}0.04$, $t_{953}$ = 2.42, p = 0.016).

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (S2 Fig; Effect of the fixed factor 'Authorship position' in a linear mixed model: Cohen's $d$ = $0.065{\pm}0.02$, $t_{2024}$ = 4.28, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors' genders than the gender of the last author, i.e. the opposite of what is predicted if senior scientists are causally responsible for homophily.

## Variance in homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was $0.12{\pm}0.02$ for Urology journals and $0.03{\pm}0.01$ for Veterinary Medicine journals; \autoref{alpha_histograms}, S3 Table). However, there was no evidence for consistent differences in $\alpha'$ between disciplines: the random factor 'Discipline' explained around 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). Thus, we cannot reject the null hypothesis that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

@@

@@

![There is a weakly positive, non-linear relationship between the gender ratio of authors publishing in a journal, and the coefficient of homophily ($\alpha'$). Specifically, journals with 50% women authors or higher tended to have more same-sex coauthorships than did journals with predominantly men authors.  This relationship held whether $\alpha'$ was calculated for all authors, first authors only, or last authors only. A negative value on the x-axis denotes an excess of men authors, a positive value denotes an excess of women authors, and zero denotes gender parity. The lines were fitted using generalised additive models with the smoothing parameter $k$ set to 3. \label{alpha_gender_ratio}](../figures/figure3.pdf)

@@

![Journal impact factor (expressed relative to the average for the discipline) is negatively correlated with $\alpha'$. The relationship is noisy ($R^2$ = `r R2`), but the results suggest that journals with strong homophily tend to have lower impact factors than journals with weak homophily in the same discipline. \label{impact_factor}](../figures/figure4.pdf)

## Analysis accounting for differences in author gender ratio between countries
When we restricted the analysis by country, we observed statistically significant homophily for 72 of the 325 journal-country combinations tested (64 unique journals and 18 unique countries), and no significant evidence of heterophily (Figures S3-S4). Additionally, the values of $\alpha'$ calculated for each journal-country combination were mostly very similar to the $\alpha'$ values calculated for the journal as a whole (i.e. when pooling papers from different countries); on average, $\alpha'$ was only inflated in the main (non-country specific) analysis by 0.002 (S5 Fig). These results suggest that our findings of widespread homophily in the main analysis were not driven solely by a Wahlund effect resulting from gender differences between countries.

@@

![When there is a difference in gender ratio between early-career and established researchers, and collaboration is non-random with respect to career stage, the null expectation for $\alpha$ deviates from zero. An excess of collaborations between career stages gives the appearance of gender heterophily (lower rows, blue areas), while an excess of within-career stage collaborations produced apparent gender homophily (upper rows, red areas). However, the conditions required for strong gender homophily are quite restrictive, making it unlikely that this issue explains all of the homophily observed in Figure 2. Contour lines denote increments of 0.1. \label{simulation}](../figures/figure5 - Simulation.pdf)

@@

@@

As expected, $\alpha'$ was usually almost identical to $\alpha$ (S6 Fig), but $\alpha$ was downwardly biased relative to $\alpha'$ for small datasets (S7 Fig). Additionally, the correlation between $\alpha'$ and sample size was negligible ($R^2 < 0.01$), suggesting that our calculation of $\alpha'$ effectively removed the dependence of $\alpha$ on sample size. We therefore used $\alpha'$ in all analyses.

@@

## Minimising the Wahlund effect: author country of affiliation
A Wahlund effect could arise even if one calculates $\alpha'$ for a single discipline and time period, because of variation in the gender ratio of researchers from different countries. For example, Holman et al. [-@Holman_2018] showed that PubMed-indexed authors based in Serbia are more than twice as likely to be women as are authors based in Japan. Therefore, a dataset containing a mix of papers from teams of authors based in these two countries would contain an excess of same-sex coauthorships, even if collaboration were random with respect to gender within each country. 
To address this issue, we also analysed every combination of journal and author country of affiliation for which we had enough data (i.e. 50 or more papers published in 2015-16). For simplicity, we restricted the dataset to only include papers for which Holman et al. [-@Holman_2018] had identified the country of affiliation for all authors on the paper, and all authors shared the same country of affiliation. Restricting the dataset in this fashion produced enough data to measure $\alpha'$ for 325 combinations of journal and country (median: 70 papers and 273 authors per combination).

@@

@@

## Data availability and reproducibility
The Online Supplementary Material contains R scripts used to produce all results, figures and tables. The input data from Holman et al. [-@Holman_2018] is archived at https://osf.io/bt9ya/ along with the code used to obtain it.

@@

\newpage
# Supporting information
S1 Fig.  There is a very strong correlation between the values of $\alpha$ and $\alpha'$ calculated for each journal, though in a handful of cases the difference is considerable. The deviation between $\alpha$ and $\alpha'$ is greatest for journals for which there is a small sample size (see S2 Fig).

S2 Fig. For journals for which we recovered a small number of papers (<100), the unadjusted metric $\alpha$ was downwardly biased. This fits our expectations: because authors cannot be their own co-authors, small datasets will tend to produce negative estimates of $\alpha$ even if authors assort randomly with respect to gender (see main text). This suggests that $\alpha'$ is a more useful measure of homophily and heterophily, especially for small samples.

S3 Fig. Histogram showing the distribution of differences in $\alpha'$ between the 2015-16 and 2005-6 samples, where positive numbers indicate an increase in $\alpha'$ with time. The mean is slightly positive (namely 0.004), indicating a mild increase in average $\alpha'$ with time.

S4 Fig. Histogram showing the difference between $\alpha'$ calculated for first and last authors. Positive values mean that $\alpha'$ was higher when calculated for first authors, and negative values mean $\alpha'$ was higher when calculated for last authors. The mean is very slightly higher than zero, indicating that $\alpha'$ tends to be higher for first authors.

S5 Fig. Histogram of $\alpha'$ for 325 unique combinations of journal and country, using data from August 2015 - August 2016. The white areas denote combinations for which $\alpha'$ differs significantly from zero (p < 0.05, following false discovery rate correction).

S6 Fig. Plot showing the 68 combinations of journal and author country of affiliation for which $\alpha'$ is significantly higher than expected.

S7 Fig. Histogram showing the estimated degree to which $\alpha'$ is inflated by inter-country differences in author gender ratio, across the 285 journals for which we had adequate data after restricting the analysis by country. The average inflation in $\alpha'$ is negligible, suggesting that Wahlund effects resulting from inter-country differences have a neglible effect on our estimates of gender homophily.

S1 Table. Sample sizes for the two datasets, which comprise papers published in the timeframes August 2005 - August 2006, and August 2015 - August 2016.

@@

S3 Table. The table shows similar information to S3 Fig, namely the distribution of the $\alpha'$ values across journals, split by the research discipline. The gender ratio column shows the percentage of women authors in the sample used to calculate $\alpha'$, across all authorship positions. In the last two columns, the numbers outside parentheses give the number of journals that deviate statistically significantly from zero, while the numbers inside parentheses give the number that remain significant after false discovery rate correction.

S1 Data: This spreadsheet shows the $\alpha$ values calculated for each journal, in the 2005 and 2015 samples, and for each type of author (all authors, first authors, and last authors). The tables gives the impact factor of each journal, the sample size, $\alpha$ and $\alpha'$ and their 95% CIs, and the p-value from a 2-tailed test evaluating the null hypothesis that $\alpha$ is zero (both raw and FDR-corrected p-values are shown).

>>>>

---
title: "Researchers preferentially collaborate with same-gendered colleagues across the life sciences"
author:
- Luke Holman* and Claire Morandin\textsuperscript{§}
- '*\textit{luke.holman@unimelb.edu.au}'
- \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}
bibliography: references.bib
output:
  pdf_document: default
  html_document: default
  word_document: default
fontsize: 12pt
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Gender and coauthorship} \usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers
csl: plos.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious
  biases, and often feel excluded from formal and informal opportunities for research
  collaboration. In addition to ensuring fairness and helping to redress gender imbalance
  in the academic workforce, increasing women's access to collaboration could help
  scientific progress by drawing on more of the available human capital. Here, we
  test whether researchers preferentially collaborate with same-gendered colleagues,
  using more stringent methods and a larger dataset than in past work. Our results
  reaffirm that researchers co-publish with colleagues of the same gender, and show
  that this 'gender homophily' is slightly stronger today than it was 10 years ago.
  Contrary to our expectations, we found no evidence that homophily is driven mostly
  by senior academics, and no evidence that homophily is stronger in fields where
  women are in the minority. Interestingly, journals with a high impact factor for their 
  discipline tended to have comparatively low homophily, as predicted if mixed-gender teams 
  produce better research. We discuss potential causes of gender homophily in science. 
  \vspace{5mm} \par\noindent
  \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining,
  Women in STEM. 
---

@@

# Load some little files with numbers to be inserted into the MS
# sample.size <- read.csv("sample.size.table.csv", stringsAsFactors = FALSE)
# sample.size[,2] <- prettyNum(sample.size[,2], big.mark = ",")
# sample.size[,3] <- prettyNum(sample.size[,3], big.mark = ",")
# names(sample.size)[2:3] <- c("n2005", "n2015")

@@

R2 <- readRDS("R2_for_figure5.RData") # For Fig 4
```

@@

Because publishing, networking and collaboration are instrumental to scientific productivity and academic career advancement [@Lee_2005; @Wuchty_2007; @Abramo_2009; @Lariviere_2015], dozens of studies have tested for gender differences in these areas [e.g. @Long_1992; @Bozeman_2011; @Abramo_2013; @Badar_2013; studies listed in Table S1 of @Holman_2018]. For example, studies have concluded that women tend to be less involved in international collaboration [@Lewison_2001; @Webster_2001; @Bozeman_2004; @Lariviere_2011; @Abramo_2013], collaborate less within their own university departments [@Webster_2001], have less prestigious collaborations [@Long_1990], and fewer collaborations in total [@Fuchs_2001]. These gender differences in collaboration practice presumably have multiple causes, which might include implicit and explicit gender bias [@Moss_2012], differential family obligations [@Reskin_1978; @Long_1990; @Wright_2003], gender differences in confidence or self-esteem [@bleidorn2016age], concerns relating to sexual harassment [@jagsi_2016], and unequal access to conferences [@Martin_2014] or travel funds [@Bozeman_2004].

A high, steadily increasing proportion of research papers is written by more than one author [@West_2013], making collaboration a key predictor of publication output, and thus of career prospects [@Tower_2007; @Jordan_2008]. Additionally, empirical studies imply that mixed-gender or otherwise diverse teams produce better outputs on collaborative tasks than less diverse teams [@Britton_2000; @Reagans_2001; @Hong_2004; @Whittington_2008; @Bear_2011; @Herrera_2012; @Campbell_2013]. For reasons such as these, multiple studies have examined the author lists of published research articles in order to test for gender differences in collaboration frequency or pattern. To our knowledge, all such studies imply that men co-publish with other men, and women with women, more often than expected if collaborators assort randomly with respect to gender [@Ferber_1980; @McDowell_1992; @Ghiasi_2015; @Crow_2015; @Fahmy_2017; @Zettler_2017; @Jadidi_2017; @Teele_2017; @Araujo_2017a; @Araujo_2017b]. This pattern of assortative publishing is often termed 'gender homophily'.

@@

![The Wahlund effect can make it appear as if authors prefer to publish with same-gendered colleagues, even if no such preference exists. Here, coloured circles represent male and female authors, and coauthors are linked with lines. Across the whole set of ten papers, there is an apparent excess of same-gender collaborations: there are six same-gender papers and only four mixed-gender papers, which is fewer than the $10\times2\times0.5\times0.5 = 5$ mixed-gender papers expected under the null hypothesis that authors assort randomly with respect to gender. However, within each subset, there is no evidence that authors prefer to publish with same-gendered individuals (if anything, this small dataset suggests gender heterophily). The Wahlund effect will tend to inflate the frequency of same-sex coauthorships whenever the data is composed of two or more disconnected subsets of literature with different author gender ratios; these subsets could be research disciplines, older versus newer papers, or papers from authors in different countries. \label{wahlund_plot}](../figures/Fig1.pdf)

@@

# Results
## Gender homophily by discipline, time period, and authorship position
\autoref{alpha_histograms} shows the distribution of $\alpha'$ estimates in 2015-2016 across all journals for which we recovered sufficient data, when $\alpha'$ was calculated for all authors, first authors only, or last authors only. The great majority of journals had $\alpha' > 0$, and for many of these the FDR-corrected p-values suggested that $\alpha'$ was significantly greater than zero (1469/2077 journals were significant in 2015-16, and 404/1192 in 2005-6; S1 Data). Only 2/2077 journals had statistically significantly heterophily (i.e. $\alpha' < 0$) in 2015-16, and 1/1192 in 2005-6 (S2 Table). The remaining 606 or 787 journals (in 2015 and 2005 respectively) had a value of $\alpha'$ not significantly different from zero, consistent with the null hypothesis of random assortment with respect to gender. We also confirmed that the majority of papers had multiple authors, in most journals (S2 Data) and disciplines (S3 Data, S1 Fig).

![Of the `r n.journals` journals for which we had adequate data in 2015-2016, `r n.sig.homo` showed statistically significant evidence of gender homophily (denoted by $\alpha' > 0$), and `r n.sig.hetero` showed statistically significant evidence of heterophily ($\alpha' < 0$), after false discovery rate correction. The white area shows the number of journals for which homophily was significantly stronger than expected under the null hypothesis (p < 0.05), while the blue area shows all the remainder. Patterns were similar whether $\alpha'$ was calculated for all authors, for first authors only, or for last authors only. \label{alpha_histograms}](../figures/Fig2.pdf)

$\alpha'$ was significantly higher in the literature sample from 2015-16 relative to 2005-6, though the difference in means was small (S2 Fig; Effect of the fixed factor 'Time period' in a linear mixed model of the data for all author positions: Cohen's $d$ = $0.091{\pm}0.04$, $t_{953}$ = 2.42, p = 0.016).

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (S3 Fig; Effect of the fixed factor 'Authorship position' in a linear mixed model: Cohen's $d$ = $0.065{\pm}0.02$, $t_{2024}$ = 4.28, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors' genders than the gender of the last author, i.e. the opposite of what is predicted if senior scientists are causally responsible for homophily.

## Variance in homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was $0.12{\pm}0.02$ for Urology journals and $0.03{\pm}0.01$ for Veterinary Medicine journals; \autoref{alpha_histograms}, S4 Data). However, there was no evidence for consistent differences in $\alpha'$ between disciplines: the random factor 'Discipline' explained around 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). Thus, we cannot reject the null hypothesis that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

@@

## Relationship between gender homophily and number of authors
Papers with two authors had significantly lower (but still positive) $\alpha'$ values than papers with more than two authors, on average (\autoref{author_number}; model output in Online Supplmentary Material). Papers with 3, 4 or 5+ authors had essentially identical average $\alpha'$ values. The variance in $\alpha'$ across journals was also a little higher for 2-authors papers compared to the remainder (\autoref{author_number}), though part of this variance is due to the reduced sample size (in terms of number of authors) for the 2-author papers.

![The coefficient of homophily ($\alpha'$) was slightly less positive when calculated for two-author papers only, relative to papers with longer author lists. The individual points, whose distribution is summarised by the violin plots, correspond to individual journals. The larger white points show the mean for each group (and its 95% CIs), as calculated by a Bayesian meta-regression model accounting for repeated measures of $\alpha'$ within journals, as well as the precision with which $\alpha'$ was estimated. \label{author_number}](../figures/Fig3.pdf)

@@

![There is a weakly positive, non-linear relationship between the gender ratio of authors publishing in a journal, and the coefficient of homophily ($\alpha'$). Specifically, journals with 50% women authors or higher tended to have more same-sex coauthorships than did journals with predominantly men authors.  This relationship held whether $\alpha'$ was calculated for all authors, first authors only, or last authors only. A negative value on the x-axis denotes an excess of men authors, a positive value denotes an excess of women authors, and zero denotes gender parity. The lines were fitted using generalised additive models with the smoothing parameter $k$ set to 3. \label{alpha_gender_ratio}](../figures/Fig4.pdf)

@@

![Journal impact factor (expressed relative to the average for the discipline) is negatively correlated with $\alpha'$. The relationship is noisy ($R^2$ = `r R2`), but the results suggest that journals with strong homophily tend to have lower impact factors than journals with weak homophily in the same discipline. \label{impact_factor}](../figures/Fig5.pdf)

## Analysis accounting for differences in author gender ratio between countries
When we restricted the analysis by country, we observed statistically significant homophily for 72 of the 325 journal-country combinations tested (64 unique journals and 18 unique countries), and no significant evidence of heterophily (S4-S5 Fig). Additionally, the values of $\alpha'$ calculated for each journal-country combination were mostly very similar to the $\alpha'$ values calculated for the journal as a whole (i.e. when pooling papers from different countries); on average, $\alpha'$ was only inflated in the main (non-country specific) analysis by _c_. 0.002 (S6 Fig). These results suggest that our findings of widespread homophily in the main analysis were not driven solely by a Wahlund effect resulting from gender differences between countries.

@@

![When there is a difference in gender ratio between early-career and established researchers, and collaboration is non-random with respect to career stage, the null expectation for $\alpha$ deviates from zero. An excess of collaborations between career stages gives the appearance of gender heterophily (lower rows, blue areas), while an excess of within-career stage collaborations produced apparent gender homophily (upper rows, red areas). However, the conditions required for strong gender homophily are quite restrictive, making it unlikely that this issue explains all of the homophily observed in Figure 2. Contour lines denote increments of 0.1. \label{simulation}](../figures/Fig6_inkscape.pdf)

@@

We also calculated $\alpha$ for papers with 2, 3, 4 or ${\ge}5$ authors, for all journals that had at least 50 suitable papers from 2015-2016 with the specified author list length.

@@

As expected, $\alpha'$ was usually almost identical to $\alpha$ (S7 Fig), but $\alpha$ was downwardly biased relative to $\alpha'$ for small datasets (S8 Fig). Additionally, the correlation between $\alpha'$ and sample size was negligible ($R^2 < 0.01$), suggesting that our calculation of $\alpha'$ effectively removed the dependence of $\alpha$ on sample size. We therefore used $\alpha'$ in all analyses.

@@

## Minimising the Wahlund effect: author country of affiliation
A Wahlund effect could arise even if one calculates $\alpha'$ for a single discipline and time period, because of variation in the gender ratio of researchers from different countries. For example, Holman et al. [-@Holman_2018] showed that PubMed-indexed authors based in Serbia are more than twice as likely to be women as are authors based in Japan. Therefore, a dataset containing a mix of papers from teams of authors based in these two countries would contain an excess of same-sex coauthorships, even if collaboration were random with respect to gender within each country.

To address this issue, we also analysed every combination of journal and author country of affiliation for which we had enough data (i.e. 50 or more papers published in 2015-16). For simplicity, we restricted the dataset to only include papers for which Holman et al. [-@Holman_2018] had identified the country of affiliation for all authors on the paper, and all authors shared the same country of affiliation. Restricting the dataset in this fashion produced enough data to measure $\alpha'$ for 325 combinations of journal and country (median: 70 papers and 273 authors per combination).

@@

To model the relationship between $\alpha'$ and the number of authors on the paper, we used a meta-regression model implemented in the R package `brms` [@burkner_2016]. The model incorporated the standard error associated with easch estimate of $\alpha'$, had author number as a fixed effect, and journal as a random intercept (to control for repeated measures of each journal). We also fit a random slope of author number within journal, thereby allowing the response to author number to vary between journals. We used the default (weak) priors. The full output of this model can be viewed in the Online Supplementary Material.

@@

## Data availability and reproducibility
The Online Supplementary Material contains R scripts used to produce all results, figures and tables; it can be viewed online at https://lukeholman.github.io/genderHomophily/. The input data from Holman et al. [-@Holman_2018] is archived at https://osf.io/bt9ya/.

@@

\newpage
# Supporting information

## Supplementary figures
S1 Fig. Plot showing the percentage of papers that have 1, 2, 3, 4, or ${\ge}5$ authors for each discipline in the dataset of Holman et al. (2018). This information can also be found in S3 Data.

S2 Fig. Histogram showing the distribution of differences in $\alpha'$ between the 2015-16 and 2005-6 samples, where positive numbers indicate an increase in $\alpha'$ with time. The mean is slightly positive (i.e. 0.004), indicating a mild increase in average $\alpha'$ with time.

S3 Fig. Histogram showing the difference between $\alpha'$ calculated for first and last authors. Positive values mean that $\alpha'$ was higher when calculated for first authors, and negative values mean $\alpha'$ was higher when calculated for last authors. The mean is very slightly higher than zero, indicating that $\alpha'$ tends to be higher for first authors.

S4 Fig. Histogram of $\alpha'$ for 325 unique combinations of journal and country, using data from August 2015 - August 2016. The white areas denote combinations for which $\alpha'$ differs significantly from zero (p < 0.05, following false discovery rate correction).

S5 Fig. Plot showing the 68 combinations of journal and author country of affiliation for which $\alpha'$ is significantly higher than expected.

S6 Fig. Histogram showing the estimated degree to which $\alpha'$ is inflated by inter-country differences in author gender ratio, across the 285 journals for which we had adequate data after restricting the analysis by country. The average inflation in $\alpha'$ is negligible, suggesting that Wahlund effects resulting from inter-country differences have a neglible effect on our estimates of gender homophily.

S7 Fig. There is a very strong correlation between the values of $\alpha$ and $\alpha'$ calculated for each journal, though in a handful of cases the difference is considerable. The deviation between $\alpha$ and $\alpha'$ is greatest for journals for which there is a small sample size (see S8 Fig).

S8 Fig. For journals for which we recovered a small number of papers (<100), the unadjusted metric $\alpha$ was downwardly biased. This fits our expectations: because authors cannot be their own co-authors, small datasets will tend to produce negative estimates of $\alpha$ even if authors assort randomly with respect to gender (see main text). This suggests that $\alpha'$ is a more useful measure of homophily and heterophily, especially for small samples.

## Supplementary tables
S1 Table. Sample sizes for the two datasets, which comprise papers published in the timeframes August 2005 - August 2006, and August 2015 - August 2016.

@@

## Supplementary datasets
S1 Data: This spreadsheet shows the $\alpha$ values calculated for each journal, in the 2005 and 2015 samples, and for each type of author (all authors, first authors, and last authors). The tables gives the impact factor of each journal, the sample size, $\alpha$ and $\alpha'$ and their 95% CIs, and the p-value from a 2-tailed test evaluating the null hypothesis that $\alpha$ is zero (both raw and FDR-corrected p-values are shown).

S2 Data: This file gives the number and percentage of paper that have 1, 2, 3, 4, or ${\ge}5$ authors for each _journal_ in the dataset of Holman et al. (2018) _PLoS Biology_. Note that the sample sizes include papers for which the gender of one or more authors was not determined by Holman et al.

S3 Data: This file gives the number and percentage of paper that have 1, 2, 3, 4, or ${\ge}5$ authors for each _discipline_ in the dataset of Holman et al. (2018) _PLoS Biology_. Note that the sample sizes include papers for which the gender of one or more authors was not determined by Holman et al.

S4 Data. The table shows the distribution of the $\alpha'$ values across journals, split by the research discipline. The gender ratio column shows the percentage of women authors in the sample used to calculate $\alpha'$, across all authorship positions. In the last two columns, the numbers outside parentheses give the number of journals that deviate statistically significantly from zero, while the numbers inside parentheses give the number that remain significant after false discovery rate correction.
